---
title: "NicheTypeR: context-aware annotation auditing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{NicheTypeR application-note workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## Overview

`NicheTypeR` is an interpretable annotation audit framework for spatial
transcriptomics. It does not try to replace expert review or atlas mapping.
Instead, it decomposes each candidate cell identity into marker, reference,
pathway or metabolic, spatial neighborhood, and ligand-receptor evidence.

The primary output is a call table with confidence, margin, and evidence
conflict reasons. The primary validation unit should be a held-out FOV, slide,
donor, or study, not a random cell split.

## Minimal workflow

```{r}
library(NicheTypeR)

toy <- simulate_nichetype_data(n_per_type = 40, seed = 1)
edges <- build_niche_graph(toy$coords, k = 8)

marker <- score_markers(toy$expr, toy$marker_db)$scores
pathway <- score_pathway_context(
  toy$expr,
  toy$pathway_sets,
  toy$label_pathways
)$scores
niche <- score_neighborhood(edges, marker, toy$niche_db)
lr <- score_lr_context(toy$expr, edges, marker, toy$lr_db)

fit <- infer_context_type(
  list(
    marker = marker,
    pathway = pathway,
    neighborhood = niche,
    ligand_receptor = lr
  ),
  weights = c(marker = 1.5, pathway = 0.5, neighborhood = 0.1, ligand_receptor = 0.5)
)

head(fit$calls)
```

## Reviewer-grade validation

When labels are available, keep the validation split spatially blocked and tune
weights only on training blocks.

```{r}
truth <- toy$metadata$true_label
names(truth) <- rownames(toy$metadata)

groups <- toy$metadata$true_label
names(groups) <- rownames(toy$metadata)
folds <- make_group_folds(truth, groups, k = 3)

learned <- learn_niche_prior(
  edges = edges,
  labels = truth,
  train_cells = folds[[1]]$train,
  candidate_labels = colnames(marker)
)

learned_niche <- score_neighborhood(edges, marker, learned$niche_db)

null_edges <- spatial_null_edges(
  edges,
  data.frame(cell_id = rownames(toy$metadata), group = groups),
  group_col = "group"
)

random_niche <- score_neighborhood(null_edges, marker, learned$niche_db)
permuted_niche <- score_neighborhood(edges, marker, permute_niche_prior(learned$niche_db))
```

## Null-corrected context-specific score

The stricter context-specific score subtracts null explanations from the
observed learned-neighborhood score before evidence fusion.

```{r}
context_specific <- score_context_specific_neighborhood(
  edges = edges,
  prior_scores = marker,
  niche_db = learned$niche_db,
  null_edges = null_edges,
  null_niche_db = permute_niche_prior(learned$niche_db),
  statistic = "residual"
)

fit_context_specific <- infer_context_type(
  list(marker = marker, context_specific = context_specific$scores),
  weights = c(marker = 1.5, context_specific = 0.1)
)
```

## Simple spatial smoothing baseline

Use `score_spatial_smoothing()` as a deliberately simple comparator. If this
baseline matches or exceeds a curated or learned niche prior, the manuscript
should not claim a specific biological niche effect.

```{r}
smoothing <- score_spatial_smoothing(edges, marker, alpha = 0.5)
neighbor_vote <- score_neighbor_majority(edges, marker)

tuned <- tune_evidence_weights(
  list(marker = marker, pathway = pathway, spatial_smoothing = smoothing, ligand_receptor = lr),
  truth = truth
)

tuned$summary[1:5, ]
```

## Interpreting conflicts

```{r}
conflicts <- summarize_evidence_conflicts(fit)
head(conflicts, 20)

clusters <- paste0("cluster_", toy$metadata$true_label)
names(clusters) <- rownames(toy$metadata)
summarize_cluster_calls(fit, clusters)
```

## Reporting standard

For a credible application note, report:

- marker-only and marker + pathway + ligand-receptor baselines
- simple spatial smoothing and random-graph controls
- learned or curated niche priors and permuted-prior controls
- held-out FOV, slide, donor, or study validation
- paired exact tests and bootstrap confidence intervals
- conflict-rate and margin summaries, not only accuracy
